End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}

### 2020-01-22 doc/2020-CAV/paper.tex
Draft paper updates

\begin{document}
\lstset{language=Ada}
\lstset{basicstyle=\ttfamily}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\title{End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
%
\titlerunning{End-to-End Verification of GR(1) Designs in SPARK}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
Joffrey Heugenot\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{L. Humphrey et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
\email{laura.humphrey@us.af.mil} \and
LinQuest Corp., Beavercreek, OH 45431, USA \\
\email{james.hamil@ctr.us.af.mil}  \and
AdaCore, F-75009 Paris, France \\
 \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.  
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising their original GR(1) specifications, namely system initial and transition properties. 
We also discuss leasons learned about how to best encode controllers in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

Several tools for GR(1) synthesis are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to easily tailor synthesis algorithms for specific applications,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable sofware implementations in a variety of programming languages.

Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
tool implementation errors could still result in errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
annotating the code, e.g. with assertions and loop invariants. 
However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifiations, 
i.e. system initial and transition properties, for moderately-sized controllers.
In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from a variety of sources.
In \sectref{sec:discussion}, we discuss possible approaches for proving liveness properties as well as design choice alternatives, 
e.g. for dealing with environment specifications.
We end with concluding remarks in \sectref{sec:conclusions}.

@@

\sloppy{
In SPARK, subunits consist of a specification and a body. 
For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
In the public part of the specification (line 2), \lstinline{Controller} is declared as a private type so that the user cannot arbitrarily manipulate its state 
(possibly violating the controller's GR(1) specification). 
In the private part of the specification (lines 34-39), \lstinline{Controller} is a record that stores the controller's internal state number, 
current input value(s), and the current output value(s). 
Its initial state is always the last possible state number. 
Each input and output is of type \lstinline{Boolean}.
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-7) or \lstinline{Environment}, respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since controllers are Moore machines, \lstinline{Controller} includes a type invariant (lines 38-39) that specifies that for each internal state, 
there is exactly one set of input values that brought the controller into that state and exactly one set of output values produced in that state. 
Functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} encode lookup tables specifying the input and output value(s) for each state, respectively.
Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they used for proof purposes only, 
i.e. they will create verification conditions related to type \lstinline{Controller} but are not executable. 
The logic for these functions is given in the body (not shown here), but for instance in state 0, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
since \lstinline{tick = False} as the system enters state 0, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
since $red$ is True and $yellow$ and $green$ are false as the system enters state 0.
 }

The public function \lstinline{Is_Init} (line 9) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
The public function \lstinline{Env_Init} (line 10) checks whether input(s) satisfy $\varphi_e^i$. 
It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
because all terms needed to define it are visible in the public part of the specification. 
In this example, since $\varphi_e^i = \top$ as pulled from the Salty specification, this function always simply returns \lstinline{True}. 
The public function \lstinline{Sys_Init} (lines 11-12) checks whether outputs(s) satisfy $\varphi_s^i$. 
It is implemented as an expression function for the same reason.  
As pulled from the Salty specification, $\varphi^s_i = red \land \lnot yellow \land \lnot green$, so this function returns the value of
the expression \lstinline{S.red and not S.yellow and not S.green}.
But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost}, 
meaning it is ``ghost code'' used only for proof and cannot be executed. 
\lstinline{Is_Init} and \lstinline{Env_Init} are used for proof and are also executable. 
We chose to make these executable for reasons related to the meaning of GR(1) specifications. 
Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
If $\varphi_e$ ever becomes false, i.e. if the environment produces input value(s) that violate $\varphi_e$, then the specification as whole 
is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
In theory, the system in this case could produce arbitrary outputs and still satisy the overall specification. 
In practice, we believe a user would generally want to know that the environment violated its specification, 
so that the user could either choose the system output value(s) explicitly or fall back to some other error handling routine. 
Therefore, a user needs to be able to check inputs with \lstinline{Env_Init} if \lstinline{Is_Init} returns true, 
which is why both are executable.
Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 14-17) check whether the next set of input value(s) and output value(s) 
satisfy $\varphi_e^t$ and $\varphi_s^t$, respectively. 
For the same reasons as above, \lstinline{Env_Trans} is executable but \lstinline{Sys_Trans} is ghost code. 
Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initial state, 
since $\varphi_e^t$ can depend on both the current and next set of input value(s). 
This precondition is not necessary for \lstinline{Sys_Trans}, since it is not executable, and it will only be used to 
assert properties in cases in which the controller is not in its initial state. 
As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is pulled from the Salty specification and 
implemented in the body (not shown), since they make use of input and output values stored in the \lstinline{Controller}, whose fields are private.

@@

A fragment of the body of \lstinline{Move} is shown in \figref{fig:moveBody}. 
Note that there are cases that can lead to \lstinline{Program_Error}. 
This is because we programmatically use \lstinline{others} to cover all possible input combinations 
that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
In the traffic light example, they are unnecessary because all possible combinations of input values are 
allowed out of each state. 
In any case, SPARK will prove that that these cases are not reachable.

@@

To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publically available for download. 
These examples -- including GR(1) specifications, sythesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section.

@@

>>>>

\begin{document}
\lstset{language=Ada}
\lstset{basicstyle=\ttfamily}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TITLE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\title{End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK\thanks{Supported by AFRL contract FA8650-16-C-2642 and AFOSR grant RQCOR20-35.}}
%
\titlerunning{End-to-End Verification of GR(1) Designs in SPARK}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Laura Humphrey\inst{1}\orcidID{0000-1111-2222-3333} \and
James Hamil\inst{2}\orcidID{1111-2222-3333-4444} \and
Joffrey Heugenot\inst{3}\orcidID{2222--3333-4444-5555}}
%
\authorrunning{L. Humphrey et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
\email{laura.humphrey@us.af.mil} \and
LinQuest Corp., Beavercreek, OH 45431, USA \\
\email{james.hamil.ctr@us.af.mil}  \and
AdaCore, F-75009 Paris, France \\
 \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.  
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising their original GR(1) specifications, namely system initial and transition properties. 
We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

Several tools for GR(1) synthesis are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and can synthesize designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to easily tailor synthesis algorithms for specific applications,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable software implementations in a variety of programming languages.

Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,'' 
tool implementation errors could still result in errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications. 
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK. 
SPARK is both a programming language with a specification language and associated verification toolset \cite{hoang2015spark}. 
Though SPARK aims to perform fully automated verification, often the user must help guide the underlying provers by
annotating the code, e.g. with assertions and loop invariants. 
However, since controller designs synthesized from GR(1) specifications follow a regular structure, it was our hope that we could 
synthesize both the controller logic and the annotations necessary to automatically prove that controllers meet their functional specifications in SPARK. 
To date, we have largely achieved this goal for a subset of properties comprising GR(1) specifications, 
i.e. system initial and transition properties, for moderately-sized controllers.
In what follows, in \sectref{sec:implementation} we discuss the SPARK implementation. 
In \sectref{sec:caseStudies}, we demonstrate the efficacy of our approach on examples pulled from a variety of sources.
In \sectref{sec:discussion}, we discuss possible approaches for proving liveness properties as well as design choice alternatives, 
e.g. for dealing with environment specifications.
We end with concluding remarks in \sectref{sec:conclusions}.

@@

\sloppy{
In SPARK, subunits consist of a specification and a body. 
For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine suitable for representing the state of a synthesized controller. 
In the public part of the specification (line 2), \lstinline{Controller} is declared as a private type so that the user cannot arbitrarily manipulate its state 
(possibly violating the controller's GR(1) specification). 
In the private part of the specification (lines 34-39), \lstinline{Controller} is a record that stores the controller's internal state number, 
current input value(s), and the current output value(s). 
Its initial state is always the last possible state number. 
Each input and output is of type \lstinline{Boolean}.
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{System} (lines 4-7) or \lstinline{Environment}, respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since controllers are Moore machines, \lstinline{Controller} includes a type invariant (lines 38-39) that specifies that for each internal state, 
there is exactly one set of input values that brought the controller into that state and exactly one set of output values produced in that state. 
Functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} encode lookup tables specifying the input and output value(s) for each state, respectively.
Both of these are declared (lines 27-30) with aspect \lstinline{Ghost}, indicating that they used for proof purposes only, 
i.e. they will create verification conditions related to type \lstinline{Controller} but are not executable. 
The logic for these functions is given in the body (not shown here), but for instance in state 1, \lstinline{State_To_Input_Equivalence} returns \lstinline{False}, 
since \lstinline{tick = False} as the system enters state 1, and \lstinline{State_To_Input_Equivalence} returns \lstinline{System'(red => True, yellow => False, green => False)}, 
since $red$ is True and $yellow$ and $green$ are false as the system enters state 1.
 }

The public function \lstinline{Is_Init} (line 9) checks whether the controller is in its initial state, i.e. no inputs have yet been received. 
The public function \lstinline{Env_Init} (line 10) checks whether input(s) satisfy $\varphi_e^i$. 
It is implemented as an expression function, i.e. the implementation is given directly in the specification and not the body, 
because all terms needed to define it are visible in the public part of the specification. 
In this example, since $\varphi_e^i = \top$ as pulled from the Salty specification, this function always simply returns \lstinline{True}. 
The public function \lstinline{Sys_Init} (lines 11-12) checks whether outputs(s) satisfy $\varphi_s^i$. 
It is implemented as an expression function for the same reason.  
As pulled from the Salty specification, $\varphi^s_i = red \land \lnot yellow \land \lnot green$, so this function returns the value of
the expression \lstinline{S.red and not S.yellow and not S.green}.
But unlike \lstinline{Is_Init} and \lstinline{Env_Init}, it is marked with aspect \lstinline{Ghost}, 
meaning it is ``ghost code'' used only for proof and cannot be executed. 
\lstinline{Is_Init} and \lstinline{Env_Init} are used for proof and are also executable. 
We chose to make these executable for reasons related to the meaning of GR(1) specifications. 
Recall that GR(1) specifications have the form $\varphi_e \limplies \varphi_s$. 
If $\varphi_e$ ever becomes false, i.e. if the environment produces input value(s) that violate $\varphi_e$, then the specification as whole 
is satisfied regardless of whether the system produces output value(s) that satisfy $\varphi_s$. 
In theory, the system in this case could produce arbitrary outputs and still satisfy the overall specification. 
In practice, we believe a user would generally want to know that the environment violated its specification, 
so that the user could either choose the system output value(s) explicitly or fall back to some other error handling routine. 
Therefore, a user needs to be able to check inputs with \lstinline{Env_Init} if \lstinline{Is_Init} returns true, 
which is why both are executable.
Public functions \lstinline{Env_Trans} and \lstinline{Sys_Trans} (lines 14-17) check whether the next set of input value(s) and output value(s) 
satisfy $\varphi_e^t$ and $\varphi_s^t$, respectively. 
For the same reasons as above, \lstinline{Env_Trans} is executable but \lstinline{Sys_Trans} is ghost code. 
Note that \lstinline{Env_Trans} has a precondition that the controller must not be in its initial state, 
since $\varphi_e^t$ can depend on both the current and next set of input value(s). 
This precondition is not necessary for \lstinline{Sys_Trans}, since it is not executable, and it will only be used to 
assert properties in cases in which the controller is not in its initial state. 
As with \lstinline{Env_Init} and \lstinline{Sys_Init}, the logic for these functions is pulled from the Salty specification and 
implemented in the body (not shown), since they make use of input and output values stored in the \lstinline{Controller}, whose fields are private.

@@

A fragment of the body of \lstinline{Move} is shown in \figref{fig:moveBody}. 
Note that there are cases that can lead to \lstinline{Program_Error}. 
This is because we programmatically use \lstinline{others} to cover all possible input combinations 
that would not be allowed due to $\varphi_e^i$ or $\varphi_e^t$. 
This makes the code generation simpler to implement as each variable's domain does not need to be explicitly tracked and covered. 
In the traffic light example, they are unnecessary because all possible combinations of input values are 
allowed out of each state. 
In any case, SPARK will prove that these cases are not reachable.

@@

To check whether our approach has the potential to produce SPARK controllers that can be automatically verified to satisfy their original system initial and transition specifications, 
we pulled examples of GR(1) specifications from Anzu, LTLMoP, TuLiP, Slugs, Salty, all of which make their examples publicly available for download. 
These examples -- including GR(1) specifications, synthesized SPARK packages, and SPARK analysis results -- are now all available in Salty format on the 
Salty GitHub repository\footnote{https://github.com/GaloisInc/salty/}, including the traffic light example of the previous section.

@@

[Laura: Summarize paper. Highlight benefits of SPARK, especially if it revealed "fall through" case statements in some examples. Mention it was relatively easy to figure out how to structure controllers
and annotate code]

[Joffrey: Speculate on how to prove liveness, whether in SPARK and/or using other tools/approaches.]
